---
title: "Earth Model Report"
format:
  html:
    toc: true
    theme: flatly
    embed-resources: true
    mainfont: "Roboto Condensed"
    include-in-header:
      text: |
        <link href="https://fonts.googleapis.com/css2?family=Roboto+Condensed:wght@400;700&display=swap" rel="stylesheet">
        <style>body { font-family: 'Roboto Condensed', sans-serif; }</style>
  pdf:
    toc: true
    mainfont: "Roboto Condensed"
    sansfont: "Roboto Condensed"
  docx:
    toc: true
    reference-doc: reference.docx
params:
  result_file: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(earthUI)
library(ggplot2)

# Load Roboto Condensed for plots in reports
if (requireNamespace("sysfonts", quietly = TRUE) &&
    requireNamespace("showtext", quietly = TRUE)) {
  sysfonts::font_add_google("Roboto Condensed", "Roboto Condensed")
  showtext::showtext_auto()
  theme_set(theme_minimal(base_family = "Roboto Condensed"))
} else {
  theme_set(theme_minimal(base_family = "sans"))
}

result <- readRDS(params$result_file)
summary_info <- format_summary(result)
multi <- isTRUE(summary_info$multi)
targets <- result$target
gf <- list_g_functions(result)
```

## Dataset Description

| Property | Value |
|----------|-------|
| Observations | `r summary_info$n_obs` |
| Target variable(s) | `r paste(targets, collapse = ", ")` |
| Predictors used | `r paste(result$predictors, collapse = ", ")` |
| Categorical predictors | `r if (length(result$categoricals) > 0) paste(result$categoricals, collapse = ", ") else "None"` |

## Model Specification

| Parameter | Value |
|-----------|-------|
| Degree | `r result$degree` |
| Cross-validation | `r if (result$cv_enabled) "Yes (10-fold)" else "No"` |
| Number of terms | `r summary_info$n_terms` |
| Predictors in model | `r summary_info$n_predictors` |
`r if (multi) paste0("| Response variables | ", length(targets), " |")`

```{r single-response-results, results='asis', eval=!multi}
cat("\n## Results Summary\n\n")
cat("| Metric | Value |\n|--------|-------|\n")
cat(sprintf("| R² | %.4f |\n", summary_info$r_squared))
cat(sprintf("| GRSq | %.4f |\n", summary_info$grsq))
cat(sprintf("| GCV | %.4f |\n", summary_info$gcv))
cat(sprintf("| RSS | %.2f |\n", summary_info$rss))
if (result$cv_enabled && !is.na(summary_info$cv_rsq)) {
  cat(sprintf("| CV R² | %.4f |\n", summary_info$cv_rsq))
}
```

```{r single-response-equation, results='asis', eval=!multi}
cat("\n## Model Equation\n\n")
eq <- format_model_equation(result)
if (!is.null(knitr::pandoc_to()) && knitr::pandoc_to() == "docx") {
  cat("$$\n", eq$latex_word, "\n$$\n")
} else {
  cat("$$\n", eq$latex_pdf, "\n$$\n")
}
```

## Coefficients & Basis Functions

```{r coefficients}
knitr::kable(summary_info$coefficients, digits = 4)
```

## Variable Importance

```{r importance-plot, fig.width=8, fig.height=5}
plot_variable_importance(result)
```

```{r importance-table}
knitr::kable(format_variable_importance(result), digits = 4)
```

```{r single-response-gfunctions, fig.width=8, fig.height=6, results='asis', eval=!multi}
cat("\n## g-Function Contributions\n\n")
for (i in seq_len(nrow(gf))) {
  label <- gf$label[i]
  if (gf$d[i] >= 2) label <- gsub(" ", " \u00d7 ", label)
  cat(sprintf("\n### %s\n\n", label))
  if (gf$d[i] >= 2) {
    plot_g_persp(result, i)
    cat("\n\n")
    p <- plot_g_contour(result, i)
    print(p)
  } else {
    p <- plot_g_contour(result, i)
    print(p)
  }
  cat("\n\n")
}
```

## Correlation Matrix

```{r correlation-matrix, fig.width=10, fig.height=8}
plot_correlation_matrix(result)
```

```{r single-response-diagnostics, fig.width=8, fig.height=5, results='asis', eval=!multi}
cat("\n## Diagnostics\n\n")
cat("\n### Residuals vs Fitted\n\n")
print(plot_residuals(result))
cat("\n\n### Normal Q-Q Plot\n\n")
print(plot_qq(result))
cat("\n\n### Actual vs Predicted\n\n")
print(plot_actual_vs_predicted(result))
```

```{r multi-response-sections, fig.width=8, fig.height=5, results='asis', eval=multi}
eq <- format_model_equation(result)
render_eq <- function(e) {
  if (!is.null(knitr::pandoc_to()) && knitr::pandoc_to() == "docx") {
    cat("$$\n", e$latex_word, "\n$$\n")
  } else {
    cat("$$\n", e$latex_pdf, "\n$$\n")
  }
}

for (ri in seq_along(targets)) {
  tgt <- targets[ri]
  cat(sprintf("\n## Response: %s\n\n", tgt))

  # --- Summary metrics ---
  cat("\n### Summary\n\n")
  cat("| Metric | Value |\n|--------|-------|\n")
  cat(sprintf("| R² | %.4f |\n", summary_info$r_squared[ri]))
  cat(sprintf("| GRSq | %.4f |\n", summary_info$grsq[ri]))
  cat(sprintf("| GCV | %.4f |\n", summary_info$gcv[ri]))
  cat(sprintf("| RSS | %.2f |\n", summary_info$rss[ri]))
  if (result$cv_enabled && !is.na(summary_info$cv_rsq[ri])) {
    cat(sprintf("| CV R² | %.4f |\n", summary_info$cv_rsq[ri]))
  }
  cat("\n")

  # --- Equation ---
  cat("\n### Model Equation\n\n")
  render_eq(eq$equations[[ri]])
  cat("\n")

  # --- g-Function contributions ---
  if (nrow(gf) > 0) {
    cat("\n### g-Function Contributions\n\n")
    for (i in seq_len(nrow(gf))) {
      label <- gf$label[i]
      if (gf$d[i] >= 2) label <- gsub(" ", " \u00d7 ", label)
      cat(sprintf("\n#### %s\n\n", label))
      if (gf$d[i] >= 2) {
        plot_g_persp(result, i, response_idx = ri)
        cat("\n\n")
        p <- plot_g_contour(result, i, response_idx = ri)
        print(p)
      } else {
        p <- plot_g_contour(result, i, response_idx = ri)
        print(p)
      }
      cat("\n\n")
    }
  }

  # --- Diagnostics ---
  cat("\n### Diagnostics\n\n")
  cat("\n#### Residuals vs Fitted\n\n")
  print(plot_residuals(result, response_idx = ri))
  cat("\n\n#### Normal Q-Q Plot\n\n")
  print(plot_qq(result, response_idx = ri))
  cat("\n\n#### Actual vs Predicted\n\n")
  print(plot_actual_vs_predicted(result, response_idx = ri))
  cat("\n\n")
}
```

## ANOVA Decomposition

```{r anova}
knitr::kable(format_anova(result), digits = 4)
```

## Earth Output

```{r earth-output, comment=""}
model <- result$model
cat("== Model ==\n\n")
print(model)
cat("\n\n== Summary ==\n\n")
print(summary(model))
if (!is.null(model$varmod)) {
  cat("\n\n== Variance Model ==\n\n")
  print(model$varmod)
}
```

---

*Report generated by [earthUI](https://github.com/wcraytor/earthUI) on `r Sys.time()`*
