---
title: "Earth Model Report"
format:
  html:
    toc: true
    theme: flatly
    mainfont: "Roboto Condensed"
    html-math-method: katex
    include-in-header:
      text: |
        <style>
        @import url('https://fonts.googleapis.com/css2?family=Roboto+Condensed:wght@400;700&display=swap');
        body { font-family: 'Roboto Condensed', sans-serif; }
        </style>
  pdf:
    toc: true
    mainfont: "Roboto Condensed"
    sansfont: "Roboto Condensed"
    include-in-header:
      text: |
        \usepackage{unicode-math}
        % Use a Times-like serif math font (matches the RCA document) so math
        % italics such as `j` are unambiguous. The previous Roboto Condensed
        % override rendered italic `j` like an elongated `i` (sans-serif has no
        % serif/curved descender). Variable names stay in the body font because
        % they are wrapped in \text{} (which uses mainfont), not math italics.
        \setmathfont{TeX Gyre Termes Math}
        \usepackage{pdflscape}
        \usepackage{longtable}
        \usepackage{booktabs}
  docx:
    toc: true
    reference-doc: reference.docx
params:
  data_file: ""
  result_file: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Resolve assets path: explicit param wins; else fall back to a sibling
# `report_data.rds` (so a self-contained bundle is renderable directly
# via `quarto render` / `quarto preview` without external arguments).
.data_file <- if (nzchar(params$data_file) && file.exists(params$data_file)) {
                params$data_file
              } else if (file.exists("report_data.rds")) {
                "report_data.rds"
              } else {
                ""
              }

# Load pre-generated assets if available, otherwise fall back to computing
if (nzchar(.data_file)) {
  .assets <- readRDS(.data_file)
  result <- .assets$result
  summary_info <- .assets$summary_info
  gf <- .assets$gf
  multi <- .assets$multi
  targets <- .assets$targets
  .plots_dir <- file.path(dirname(.data_file), "plots")
  if (.plots_dir == "./plots" || !nzchar(dirname(.data_file)) ||
      identical(dirname(.data_file), ".")) {
    .plots_dir <- "plots"
  }
  .has_assets <- dir.exists(.plots_dir) &&
    length(list.files(.plots_dir)) > 0L
} else if (nzchar(params$result_file) && file.exists(params$result_file)) {
  # Legacy path: compute everything (slower)
  library(earthUI)
  library(ggplot2)
  if (requireNamespace("sysfonts", quietly = TRUE) &&
      requireNamespace("showtext", quietly = TRUE)) {
    if (!"Roboto Condensed" %in% sysfonts::font_families()) {
      tryCatch(
        sysfonts::font_add_google("Roboto Condensed", "Roboto Condensed"),
        error = function(e) NULL)
    }
    showtext::showtext_auto()
    theme_set(theme_minimal(base_family = "Roboto Condensed"))
  } else {
    theme_set(theme_minimal(base_family = "sans"))
  }
  result <- readRDS(params$result_file)
  summary_info <- earthUI::format_summary(result)
  gf <- earthUI::list_g_functions(result)
  multi <- isTRUE(summary_info$multi)
  targets <- result$target
  .has_assets <- FALSE
  .assets <- NULL
}

fmt <- if (!is.null(knitr::pandoc_to())) knitr::pandoc_to() else "html"
.plot_ext <- if (fmt == "latex") "pdf" else "png"

# Helper: include a pre-generated plot or generate on-the-fly
include_plot <- function(name, gen_expr, width = 8, height = 5,
                         asis = FALSE) {
  if (.has_assets) {
    f <- file.path(.plots_dir, paste0(name, ".", .plot_ext))
    if (file.exists(f)) {
      if (asis) {
        if (fmt == "latex") {
          # PDF: use relative path (lualatex resolves from working dir)
          rel <- file.path("plots", paste0(name, ".", .plot_ext))
          cat(sprintf("![](%s)\n\n", rel))
        } else {
          # HTML/Word: embed as base64 data URI so image travels with the file
          cat(sprintf("![](%s)\n\n", knitr::image_uri(f)))
        }
        return(invisible(NULL))
      } else {
        return(knitr::include_graphics(f))
      }
    }
  }
  eval(gen_expr)
}
```

## Dataset Description

| Property | Value |
|----------|-------|
| Observations | `r summary_info$n_obs` |
| Target variable(s) | `r paste(targets, collapse = ", ")` |
| Predictors used | `r paste(result$predictors, collapse = ", ")` |
| Categorical predictors | `r if (length(result$categoricals) > 0) paste(result$categoricals, collapse = ", ") else "None"` |

## Model Specification

| Parameter | Value |
|-----------|-------|
| Degree | `r result$degree` |
| Cross-validation | `r if (result$cv_enabled) "Yes (10-fold)" else "No"` |
| Number of terms | `r summary_info$n_terms` |
| Predictors in model | `r summary_info$n_predictors` |
`r if (multi) paste0("| Response variables | ", length(targets), " |")`

```{r interaction-matrix, results='asis', eval=result$degree >= 2 && !is.null(result$allowed_matrix)}
mat <- result$allowed_matrix
preds <- rownames(mat)
n <- length(preds)

# Build display matrix (Yes/No, diagonal = -)
display <- matrix("", n, n, dimnames = list(preds, preds))
for (i in seq_len(n)) {
  for (j in seq_len(n)) {
    display[i, j] <- if (i == j) "" else if (mat[i, j]) "\u2713" else "\u2014"
  }
}

# Truncate long predictor names for compact display
max_len <- if (n > 8) 12 else if (n > 5) 16 else 30
short <- ifelse(nchar(preds) > max_len,
                paste0(substr(preds, 1, max_len - 1), "\u2026"), preds)
colnames(display) <- short
rownames(display) <- short

if (fmt == "latex") {
  cat("\n\\newpage\n\\begin{landscape}\n")
  cat("\\subsection{Allowed Interactions}\n\n")
  cat("\\footnotesize\n")
  print(knitr::kable(display, format = "latex", booktabs = TRUE))
  cat("\n\\normalsize\n")
  cat("\\end{landscape}\n\\newpage\n")
} else if (fmt == "docx") {
  # Landscape, with the table rendered as raw OOXML so we can set the font
  # (Arial Narrow 8pt) and tighten the page margins to 0.5". A pandoc pipe table
  # can't carry a font; a raw <w:tbl> with explicit run properties can. Word
  # uses half-points (8pt -> w:sz 16) and twips (0.5in -> 720; landscape letter
  # page = 15840 x 12240).
  esc_xml_ <- function(x) {
    x <- gsub("&", "&amp;", x, fixed = TRUE)
    x <- gsub("<", "&lt;",  x, fixed = TRUE)
    gsub(">", "&gt;",  x, fixed = TRUE)
  }
  rpr_ <- function(bold = FALSE) paste0(
    '<w:rPr><w:rFonts w:ascii="Arial Narrow" w:hAnsi="Arial Narrow" ',
    'w:cs="Arial Narrow"/>', if (bold) '<w:b/>' else '',
    '<w:sz w:val="16"/><w:szCs w:val="16"/></w:rPr>')
  cell_ <- function(txt, bold = FALSE) paste0(
    '<w:tc><w:tcPr><w:tcMar>',
    '<w:top w:w="20" w:type="dxa"/><w:bottom w:w="20" w:type="dxa"/>',
    '<w:left w:w="40" w:type="dxa"/><w:right w:w="40" w:type="dxa"/></w:tcMar>',
    '<w:vAlign w:val="center"/></w:tcPr>',
    '<w:p><w:pPr><w:jc w:val="center"/>', rpr_(bold), '</w:pPr>',
    '<w:r>', rpr_(bold), '<w:t xml:space="preserve">', esc_xml_(txt),
    '</w:t></w:r></w:p></w:tc>')
  brd_ <- paste0('<w:tblBorders>',
    paste(sprintf('<w:%s w:val="single" w:sz="4" w:space="0" w:color="999999"/>',
      c("top", "left", "bottom", "right", "insideH", "insideV")), collapse = ""),
    '</w:tblBorders>')
  header_row <- paste0('<w:tr>', cell_("", TRUE),
    paste(vapply(short, cell_, character(1), bold = TRUE), collapse = ""), '</w:tr>')
  body_rows <- paste(vapply(seq_len(n), function(i) {
    cells <- paste(vapply(seq_len(n), function(j) {
      if (i == j) cell_("") else if (mat[i, j]) cell_("✓") else cell_("—")
    }, character(1)), collapse = "")
    paste0('<w:tr>', cell_(short[i], TRUE), cells, '</w:tr>')
  }, character(1)), collapse = "")
  tbl_ <- paste0('<w:tbl><w:tblPr><w:tblW w:w="0" w:type="auto"/>', brd_,
    '</w:tblPr>', header_row, body_rows, '</w:tbl>')

  # Portrait sectPr terminates the preceding (1-inch) report body.
  cat('\n\n```{=openxml}\n')
  cat('<w:p><w:pPr><w:sectPr><w:pgSz w:w="12240" w:h="15840"/>')
  cat('<w:pgMar w:top="1440" w:right="1440" w:bottom="1440" w:left="1440"/>')
  cat('</w:sectPr></w:pPr></w:p>\n')
  cat('```\n\n')
  cat("## Allowed Interactions\n\n")
  cat('```{=openxml}\n', tbl_, '\n```\n\n', sep = "")
  # Landscape sectPr with 0.5-inch (720 twip) margins terminates this section.
  cat('```{=openxml}\n')
  cat('<w:p><w:pPr><w:sectPr><w:pgSz w:w="15840" w:h="12240" w:orient="landscape"/>')
  cat('<w:pgMar w:top="720" w:right="720" w:bottom="720" w:left="720"/>')
  cat('</w:sectPr></w:pPr></w:p>\n')
  cat('```\n\n')
} else {
  # HTML: full-width with rotated column headers for large matrices
  cat("\n## Allowed Interactions\n\n")
  cat('<style>\n')
  cat('.eui-matrix-wrap { overflow-x: auto; margin: 1em 0; }\n')
  cat('.eui-matrix { border-collapse: collapse; font-size: 0.8em; }\n')
  cat('.eui-matrix th, .eui-matrix td { border: 1px solid #ddd; padding: 3px 6px; text-align: center; }\n')
  cat('.eui-matrix th.eui-rotate { height: 120px; white-space: nowrap; vertical-align: bottom; }\n')
  cat('.eui-matrix th.eui-rotate > div { transform: rotate(-60deg); width: 1.5em; transform-origin: bottom left; margin-left: 50%; }\n')
  cat('.eui-matrix td:first-child { text-align: left; font-weight: bold; white-space: nowrap; }\n')
  cat('.eui-matrix .eui-check { color: #27ae60; font-weight: bold; }\n')
  cat('.eui-matrix .eui-dash { color: #999; }\n')
  cat('@media print { .eui-matrix-wrap { page-break-before: always; } @page { size: landscape; } }\n')
  cat('</style>\n')
  cat('<div class="eui-matrix-wrap">\n')
  cat('<table class="eui-matrix">\n<thead><tr><th></th>')
  for (j in seq_len(n)) {
    cat(sprintf('<th class="eui-rotate"><div>%s</div></th>', preds[j]))
  }
  cat('</tr></thead>\n<tbody>\n')
  for (i in seq_len(n)) {
    cat(sprintf('<tr><td>%s</td>', preds[i]))
    for (j in seq_len(n)) {
      if (i == j) {
        cat('<td></td>')
      } else if (mat[i, j]) {
        cat('<td class="eui-check">\u2713</td>')
      } else {
        cat('<td class="eui-dash">\u2014</td>')
      }
    }
    cat('</tr>\n')
  }
  cat('</tbody></table>\n</div>\n')
}
```

```{r single-response-results, results='asis', eval=!multi}
cat("\n## Results Summary\n\n")
cat("| Metric | Value |\n|--------|-------|\n")
cat(sprintf("| R² | %.4f |\n", summary_info$r_squared))
cat(sprintf("| GRSq | %.4f |\n", summary_info$grsq))
cat(sprintf("| GCV | %.4f |\n", summary_info$gcv))
cat(sprintf("| RSS | %.2f |\n", summary_info$rss))
if (result$cv_enabled && !is.na(summary_info$cv_rsq)) {
  cat(sprintf("| CV R² | %.4f |\n", summary_info$cv_rsq))
}
```

```{r single-response-equation, results='asis', eval=!multi}
cat("\n## Model Equation\n\n")
eq <- if (!is.null(.assets)) .assets$equation else earthUI::format_model_equation(result)
if (!is.null(knitr::pandoc_to()) && knitr::pandoc_to() == "docx") {
  cat("$$\n", eq$latex_word, "\n$$\n")
} else {
  cat("$$\n", eq$latex_pdf, "\n$$\n")
}
cat("\nSee [Notation: Value-Contribution ($g$) Functions](#sec-gnotation) at",
    "the end of this report for an explanation of the",
    "${}^{f}g^{j}_{k}$ notation and its indexes.\n\n")
```

## Coefficients & Basis Functions

```{r coefficients}
knitr::kable(summary_info$coefficients, digits = 4)
```

## Variable Importance

```{r importance-plot, fig.width=8, fig.height=5}
include_plot("importance", quote(earthUI::plot_variable_importance(result)))
```

```{r importance-table}
imp <- if (!is.null(.assets)) .assets$importance else earthUI::format_variable_importance(result)
knitr::kable(imp, digits = 4)
```

```{r single-response-gfunctions, fig.width=8, fig.height=6, results='asis', eval=!multi}
cat("\n## g-Function Contributions\n\n")
for (i in seq_len(nrow(gf))) {
  label <- gf$label[i]
  if (gf$d[i] >= 2) label <- gsub(" ", " \u00d7 ", label)
  cat(sprintf("\n### %s\n\n", label))
  if (gf$d[i] >= 2) {
    include_plot(paste0("gfunc_", i, "_persp"),
                 quote(earthUI::plot_g_persp(result, i)), asis = TRUE)
    cat("\n\n")
  }
  include_plot(paste0("gfunc_", i, "_contour"),
               quote(earthUI::plot_g_contour(result, i)), asis = TRUE)
  cat("\n\n")
}
```

## Correlation Matrix

```{r correlation-matrix, fig.width=10, fig.height=8}
include_plot("correlation", quote(earthUI::plot_correlation_matrix(result)),
             width = 10, height = 8)
```

```{r single-response-diagnostics, fig.width=8, fig.height=5, results='asis', eval=!multi}
cat("\n## Diagnostics\n\n")
cat("\n### Residuals vs Fitted\n\n")
include_plot("residuals", quote(print(earthUI::plot_residuals(result))), asis = TRUE)
cat("\n\n### Normal Q-Q Plot\n\n")
include_plot("qq", quote(print(earthUI::plot_qq(result))), asis = TRUE)
cat("\n\n### Actual vs Predicted\n\n")
include_plot("actual_vs_predicted",
             quote(print(earthUI::plot_actual_vs_predicted(result))), asis = TRUE)
```

```{r multi-response-sections, fig.width=8, fig.height=5, results='asis', eval=multi}
eq <- if (!is.null(.assets)) .assets$equation else earthUI::format_model_equation(result)
render_eq <- function(e) {
  if (!is.null(knitr::pandoc_to()) && knitr::pandoc_to() == "docx") {
    cat("$$\n", e$latex_word, "\n$$\n")
  } else {
    cat("$$\n", e$latex_pdf, "\n$$\n")
  }
}

for (ri in seq_along(targets)) {
  tgt <- targets[ri]
  suffix <- paste0("_r", ri)
  cat(sprintf("\n## Response: %s\n\n", tgt))

  # --- Summary metrics ---
  cat("\n### Summary\n\n")
  cat("| Metric | Value |\n|--------|-------|\n")
  cat(sprintf("| R² | %.4f |\n", summary_info$r_squared[ri]))
  cat(sprintf("| GRSq | %.4f |\n", summary_info$grsq[ri]))
  cat(sprintf("| GCV | %.4f |\n", summary_info$gcv[ri]))
  cat(sprintf("| RSS | %.2f |\n", summary_info$rss[ri]))
  if (result$cv_enabled && !is.na(summary_info$cv_rsq[ri])) {
    cat(sprintf("| CV R² | %.4f |\n", summary_info$cv_rsq[ri]))
  }
  cat("\n")

  # --- Equation ---
  cat("\n### Model Equation\n\n")
  render_eq(eq$equations[[ri]])
  cat("\nSee [Notation: Value-Contribution ($g$) Functions](#sec-gnotation) at",
      "the end of this report for an explanation of the",
      "${}^{f}g^{j}_{k}$ notation and its indexes.\n\n")

  # --- g-Function contributions ---
  if (nrow(gf) > 0) {
    cat("\n### g-Function Contributions\n\n")
    for (i in seq_len(nrow(gf))) {
      label <- gf$label[i]
      if (gf$d[i] >= 2) label <- gsub(" ", " \u00d7 ", label)
      cat(sprintf("\n#### %s\n\n", label))
      if (gf$d[i] >= 2) {
        include_plot(paste0("gfunc_", i, suffix, "_persp"),
                     bquote(earthUI::plot_g_persp(result, .(i), response_idx = .(ri))), asis = TRUE)
        cat("\n\n")
      }
      include_plot(paste0("gfunc_", i, suffix, "_contour"),
                   bquote(p <- earthUI::plot_g_contour(result, .(i), response_idx = .(ri)); print(p)), asis = TRUE)
      cat("\n\n")
    }
  }

  # --- Diagnostics ---
  cat("\n### Diagnostics\n\n")
  cat("\n#### Residuals vs Fitted\n\n")
  include_plot(paste0("residuals", suffix),
               bquote(print(earthUI::plot_residuals(result, response_idx = .(ri)))), asis = TRUE)
  cat("\n\n#### Normal Q-Q Plot\n\n")
  include_plot(paste0("qq", suffix),
               bquote(print(earthUI::plot_qq(result, response_idx = .(ri)))), asis = TRUE)
  cat("\n\n#### Actual vs Predicted\n\n")
  include_plot(paste0("actual_vs_predicted", suffix),
               bquote(print(earthUI::plot_actual_vs_predicted(result, response_idx = .(ri)))), asis = TRUE)
  cat("\n\n")
}
```

## ANOVA Decomposition

```{r anova}
anova_df <- if (!is.null(.assets)) .assets$anova else earthUI::format_anova(result)
knitr::kable(anova_df, digits = 4)
```

## Earth Output

```{r earth-output, comment=""}
if (!is.null(.assets) && !is.null(.assets$earth_output)) {
  # Pre-captured at asset-generation time (matches the app's Earth Output tab);
  # the lean result saved for the report has no model to print/summarise.
  cat(.assets$earth_output)
} else {
  # Legacy path: full result available, render directly.
  model <- result$model
  cat("== Model ==\n\n")
  print(model)
  cat("\n\n== Summary ==\n\n")
  print(summary(model))
  if (!is.null(model$varmod)) {
    cat("\n\n== Variance Model ==\n\n")
    print(model$varmod)
  }
}
```

## Notation: Value-Contribution ($g$) Functions {#sec-gnotation .unnumbered}

```{r g-legend, results='asis'}
cat(earthUI::g_function_legend("markdown"))
```

---

*Report generated by [earthUI](https://github.com/wcraytor/earthUI) on `r Sys.time()`*
