---
title: "GAM Model Report"
format:
  html:
    toc: true
    theme: flatly
    html-math-method: katex
  pdf:
    toc: true
    papersize: letter
    include-in-header:
      text: |
        \usepackage{unicode-math}
        % Times-like serif math font (matches the RCA document and the earthUI/
        % glmnetUI reports) so math italics are unambiguous and consistent
        % across the three apps.
        \setmathfont{TeX Gyre Termes Math}
        \usepackage{longtable}
        \usepackage{booktabs}
  docx:
    toc: true
    reference-doc: reference.docx
params:
  data_file: ""
  paper_size: "letter"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
`%||%` <- function(a, b) if (is.null(a)) b else a

# Resolve the assets payload: explicit param wins; else a sibling
# report_data.rds (self-contained bundle), so `quarto render` works with no
# external arguments.
.data_file <- if (nzchar(params$data_file) && file.exists(params$data_file)) {
  params$data_file
} else if (file.exists("report_data.rds")) {
  "report_data.rds"
} else {
  ""
}

.assets <- if (nzchar(.data_file)) readRDS(.data_file) else NULL
result       <- .assets$result
summary_info <- .assets$summary_info
importance   <- .assets$importance
equation       <- .assets$equation
equation_latex <- .assets$equation_latex
equation_defs  <- .assets$equation_defs
summary_text   <- .assets$summary_text

.plots_dir <- if (nzchar(.data_file)) file.path(dirname(.data_file), "plots") else "plots"
if (!dir.exists(.plots_dir) && dir.exists("plots")) .plots_dir <- "plots"
.has_assets <- dir.exists(.plots_dir) && length(list.files(.plots_dir)) > 0L

fmt <- if (!is.null(knitr::pandoc_to())) knitr::pandoc_to() else "html"
.plot_ext <- if (fmt == "latex") "pdf" else "png"

# Enabled-interaction matrix (tensor + factor-by-smooth); shown only if any.
.imat <- .assets$interaction_matrix
.has_interactions <- !is.null(.imat) && any(.imat)

# Include a pre-generated plot if it exists. For PDF use the relative path;
# for HTML/Word embed as a data URI so the image travels with the file.
include_plot <- function(name) {
  f <- file.path(.plots_dir, paste0(name, ".", .plot_ext))
  if (!file.exists(f)) return(invisible(NULL))
  if (fmt == "latex") {
    cat(sprintf("![](%s)\n\n", file.path("plots", paste0(name, ".", .plot_ext))))
  } else {
    cat(sprintf("![](%s)\n\n", knitr::image_uri(f)))
  }
  invisible(NULL)
}
```

## Dataset & Model

| Property | Value |
|----------|-------|
| Response | `r result$response` |
| Predictors | `r paste(result$predictors, collapse = ", ")` |
| Observations | `r summary_info$n_obs` |
| Family | `r summary_info$family` |
| Smoothing method | `r summary_info$method` |
| Smooth terms | `r summary_info$n_smooths` |

## Results Summary

| Metric | Value |
|--------|-------|
| R-squared (adj) | `r sprintf("%.4f", summary_info$r_squared)` |
| Deviance explained | `r sprintf("%.1f%%", 100 * summary_info$dev_explained)` |
| `r if (!is.null(summary_info$cv_rsq)) "CV R-squared" else "CV R-squared (n/a)"` | `r if (!is.null(summary_info$cv_rsq)) sprintf("%.4f", summary_info$cv_rsq) else "—"` |
| AIC | `r sprintf("%.1f", summary_info$aic)` |
| BIC | `r if (is.finite(summary_info$bic)) sprintf("%.1f", summary_info$bic) else "—"` |

## Model Equation

```{r equation, results='asis'}
if (!is.null(equation_latex) && nzchar(equation_latex)) {
  # Display math, one term per line (aligned) so it never runs off the page.
  cat("\n$$\n", equation_latex, "\n$$\n\n", sep = "")
  if (length(equation_defs)) {
    cat("**Smooth function definitions:**\n\n")
    for (d in equation_defs) {
      cat(sprintf("- $%s$ → `%s` — %s\n", d$fn, d$label, d$desc))
    }
    cat("\n")
  }
} else {
  # Fallback: raw R formula in a code block.
  cat("```\n", equation, "\n```\n", sep = "")
}
cat("See [Notation: Value-Contribution Functions](#sec-gnotation) at the end",
    "of this report for an explanation of the $f_i$ smooth-function notation.\n\n")
```

## Smooth Terms

```{r smooth-table}
st <- summary_info$smooth_table
if (nrow(st) > 0L) {
  st$EDF <- round(st$EDF, 3); st$Ref.df <- round(st$Ref.df, 3)
  st$F <- round(st$F, 3)
  st$p_value <- format.pval(st$p_value, digits = 3, eps = 1e-4)
  knitr::kable(st)
} else cat("_No smooth terms._\n")
```

## Parametric Terms

```{r param-table}
pt <- summary_info$parametric_table
if (nrow(pt) > 0L) {
  pt$Estimate <- signif(pt$Estimate, 4); pt$Std_Error <- signif(pt$Std_Error, 4)
  pt$t_value <- round(pt$t_value, 3)
  pt$p_value <- format.pval(pt$p_value, digits = 3, eps = 1e-4)
  knitr::kable(pt)
} else cat("_No parametric terms._\n")
```

```{r enabled-interactions, results='asis', eval=.has_interactions}
# ✓ = a tensor (ti/te) or factor-by-smooth interaction term between the two
# variables was included in the GAM.
mat <- .imat
preds <- rownames(mat)
n <- length(preds)

display <- matrix("", n, n, dimnames = list(preds, preds))
for (i in seq_len(n)) {
  for (j in seq_len(n)) {
    display[i, j] <- if (i == j) "" else if (mat[i, j]) "✓" else "—"
  }
}

max_len <- if (n > 8) 12 else if (n > 5) 16 else 30
short <- ifelse(nchar(preds) > max_len,
                paste0(substr(preds, 1, max_len - 1), "…"), preds)
colnames(display) <- short
rownames(display) <- short

if (fmt == "latex") {
  cat("\n\\newpage\n\\begin{landscape}\n")
  cat("\\subsection{Enabled Interactions}\n\n")
  cat("\\footnotesize\n")
  print(knitr::kable(display, format = "latex", booktabs = TRUE))
  cat("\n\\normalsize\n")
  cat("\\end{landscape}\n\\newpage\n")
} else if (fmt == "docx") {
  esc_xml_ <- function(x) {
    x <- gsub("&", "&amp;", x, fixed = TRUE)
    x <- gsub("<", "&lt;",  x, fixed = TRUE)
    gsub(">", "&gt;",  x, fixed = TRUE)
  }
  rpr_ <- function(bold = FALSE) paste0(
    '<w:rPr><w:rFonts w:ascii="Arial Narrow" w:hAnsi="Arial Narrow" ',
    'w:cs="Arial Narrow"/>', if (bold) '<w:b/>' else '',
    '<w:sz w:val="16"/><w:szCs w:val="16"/></w:rPr>')
  cell_ <- function(txt, bold = FALSE) paste0(
    '<w:tc><w:tcPr><w:tcMar>',
    '<w:top w:w="20" w:type="dxa"/><w:bottom w:w="20" w:type="dxa"/>',
    '<w:left w:w="40" w:type="dxa"/><w:right w:w="40" w:type="dxa"/></w:tcMar>',
    '<w:vAlign w:val="center"/></w:tcPr>',
    '<w:p><w:pPr><w:jc w:val="center"/>', rpr_(bold), '</w:pPr>',
    '<w:r>', rpr_(bold), '<w:t xml:space="preserve">', esc_xml_(txt),
    '</w:t></w:r></w:p></w:tc>')
  brd_ <- paste0('<w:tblBorders>',
    paste(sprintf('<w:%s w:val="single" w:sz="4" w:space="0" w:color="999999"/>',
      c("top", "left", "bottom", "right", "insideH", "insideV")), collapse = ""),
    '</w:tblBorders>')
  header_row <- paste0('<w:tr>', cell_("", TRUE),
    paste(vapply(short, cell_, character(1), bold = TRUE), collapse = ""), '</w:tr>')
  body_rows <- paste(vapply(seq_len(n), function(i) {
    cells <- paste(vapply(seq_len(n), function(j) {
      if (i == j) cell_("") else if (mat[i, j]) cell_("✓") else cell_("—")
    }, character(1)), collapse = "")
    paste0('<w:tr>', cell_(short[i], TRUE), cells, '</w:tr>')
  }, character(1)), collapse = "")
  tbl_ <- paste0('<w:tbl><w:tblPr><w:tblW w:w="0" w:type="auto"/>', brd_,
    '</w:tblPr>', header_row, body_rows, '</w:tbl>')
  cat('\n\n```{=openxml}\n')
  cat('<w:p><w:pPr><w:sectPr><w:pgSz w:w="12240" w:h="15840"/>')
  cat('<w:pgMar w:top="1440" w:right="1440" w:bottom="1440" w:left="1440"/>')
  cat('</w:sectPr></w:pPr></w:p>\n')
  cat('```\n\n')
  cat("## Enabled Interactions\n\n")
  cat('```{=openxml}\n', tbl_, '\n```\n\n', sep = "")
  cat('```{=openxml}\n')
  cat('<w:p><w:pPr><w:sectPr><w:pgSz w:w="15840" w:h="12240" w:orient="landscape"/>')
  cat('<w:pgMar w:top="720" w:right="720" w:bottom="720" w:left="720"/>')
  cat('</w:sectPr></w:pPr></w:p>\n')
  cat('```\n\n')
} else {
  cat("\n## Enabled Interactions\n\n")
  cat('<style>\n')
  cat('.int-matrix-wrap { overflow-x: auto; margin: 1em 0; }\n')
  cat('.int-matrix { border-collapse: collapse; font-size: 0.8em; }\n')
  cat('.int-matrix th, .int-matrix td { border: 1px solid #ddd; padding: 3px 6px; text-align: center; }\n')
  cat('.int-matrix th.int-rotate { height: 120px; white-space: nowrap; vertical-align: bottom; }\n')
  cat('.int-matrix th.int-rotate > div { transform: rotate(-60deg); width: 1.5em; transform-origin: bottom left; margin-left: 50%; }\n')
  cat('.int-matrix td:first-child { text-align: left; font-weight: bold; white-space: nowrap; }\n')
  cat('.int-matrix .int-check { color: #27ae60; font-weight: bold; }\n')
  cat('.int-matrix .int-dash { color: #999; }\n')
  cat('</style>\n')
  cat('<div class="int-matrix-wrap">\n')
  cat('<table class="int-matrix">\n<thead><tr><th></th>')
  for (j in seq_len(n)) {
    cat(sprintf('<th class="int-rotate"><div>%s</div></th>', preds[j]))
  }
  cat('</tr></thead>\n<tbody>\n')
  for (i in seq_len(n)) {
    cat(sprintf('<tr><td>%s</td>', preds[i]))
    for (j in seq_len(n)) {
      if (i == j) {
        cat('<td></td>')
      } else if (mat[i, j]) {
        cat('<td class="int-check">✓</td>')
      } else {
        cat('<td class="int-dash">—</td>')
      }
    }
    cat('</tr>\n')
  }
  cat('</tbody></table>\n</div>\n')
}
```

## Variable Importance

```{r importance, results='asis'}
include_plot("importance")
```

## Smooth Effects

```{r smooths, results='asis'}
# Smooth panels are split into page-sized chunks so each renders large.
nch <- .assets$n_smooth_chunks
if (is.null(nch) || nch < 1) {
  include_plot("smooths")            # backward-compat (older bundles)
} else {
  for (i in seq_len(nch)) include_plot(paste0("smooths_", i))
}
```

```{r per-term, results='asis'}
specs <- result$smooth_specs
if (!is.null(specs)) {
  for (i in seq_along(specs)) {
    sp <- specs[[i]]
    if (identical(sp$type %||% "s", "linear")) next
    lbl <- paste(sp$vars, collapse = " × ")
    cat(sprintf("\n### %s\n\n", lbl))
    include_plot(paste0("term_", i))
  }
}
```

## Predictor Correlations

```{r correlation, results='asis'}
include_plot("correlation")
```

## Diagnostics

```{r diagnostics, results='asis'}
include_plot("diagnostics")
```

### Actual vs Predicted

```{r avp, results='asis'}
include_plot("actual_vs_predicted")
```

## mgcv Summary

```{r raw-summary, comment=""}
cat(summary_text)
```

## Notation: Value-Contribution Functions {#sec-gnotation .unnumbered}

```{r g-legend, results='asis'}
cat(earthUI::g_function_legend_mgcv("markdown"))
```

---

*Report generated by mgcvUI on `r format(Sys.time())`.*
