---
title: "glmnetUI Model Report"
format:
  html:
    toc: true
    theme: flatly
    mainfont: "Roboto Condensed"
    html-math-method: katex
    embed-resources: true
    include-in-header:
      text: |
        <style>
        @import url('https://fonts.googleapis.com/css2?family=Roboto+Condensed:wght@400;700&display=swap');
        body { font-family: 'Roboto Condensed', sans-serif; }
        </style>
  pdf:
    toc: true
    mainfont: "Roboto Condensed"
    sansfont: "Roboto Condensed"
    include-in-header:
      text: |
        \usepackage{unicode-math}
        % Use a Times-like serif math font (matches the RCA document) so math
        % italics such as `j` are unambiguous. The previous Roboto Condensed
        % override rendered italic `j` like an elongated `i` (sans-serif has no
        % serif/curved descender). Variable names stay in the body font because
        % they are wrapped in \text{} (which uses mainfont), not math italics.
        \setmathfont{TeX Gyre Termes Math}
        \usepackage{longtable}
        \usepackage{booktabs}
        \usepackage{lscape}
  docx:
    toc: true
    reference-doc: reference.docx
params:
  data_file: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

if (nzchar(params$data_file) && file.exists(params$data_file)) {
  .assets <- readRDS(params$data_file)
  .plots_dir <- file.path(dirname(params$data_file), "plots")
  .has_assets <- dir.exists(.plots_dir) &&
    length(list.files(.plots_dir)) > 0L
} else {
  stop("No data_file provided.")
}

fmt <- if (!is.null(knitr::pandoc_to())) knitr::pandoc_to() else "html"
.plot_ext <- if (fmt == "latex") "pdf" else "png"

# Enabled-interaction matrix (only shown if any pairwise interaction was used).
.imat <- .assets$interaction_matrix
.has_interactions <- !is.null(.imat) && any(.imat)

include_plot <- function(name, width = 8, height = 5) {
  if (.has_assets) {
    f <- file.path(.plots_dir, paste0(name, ".", .plot_ext))
    if (file.exists(f)) {
      # Use relative path — plots/ is next to the QMD in the render dir.
      # Absolute paths break in LaTeX (luatex prepends ./).
      rel <- file.path("plots", paste0(name, ".", .plot_ext))
      cat(sprintf("![](%s)\n\n", rel))
    }
  }
}
```

## Dataset Description

| Property | Value |
|----------|-------|
| Data file | `r .assets$data_file_name` |
| Observations | `r .assets$n_obs` |
| Response variable | `r .assets$response` |
| Predictors | `r paste(.assets$predictors, collapse = ", ")` |
| Purpose | `r .assets$purpose` |

## Model Specification

| Parameter | Value |
|-----------|-------|
| Alpha | `r .assets$alpha` |
| Lambda | `r signif(.assets$lambda, 5)` |
| Lambda selection | `r .assets$lambda_method` |
| Family | `r .assets$family` |
| Standardize | `r .assets$standardize` |
| Relaxed | `r .assets$relaxed` |
| Non-zero coefficients | `r .assets$n_nonzero` |
| Total predictors in model | `r .assets$n_predictors` |

## Results Summary

| Metric | Value |
|--------|-------|
| R² | `r sprintf("%.4f", .assets$r_squared)` |
| Adjusted R² | `r sprintf("%.4f", .assets$adj_r_squared)` |
| Generalized R² | `r sprintf("%.4f", .assets$gen_r_squared)` |
| CV R² | `r if (!is.na(.assets$cv_r_squared)) sprintf("%.4f", .assets$cv_r_squared) else "N/A"` |
| RMSE | `r sprintf("%.2f", .assets$rmse)` |
| MAE | `r sprintf("%.2f", .assets$mae)` |

## Model Equation

```{r equation, results='asis'}
# `\small` is a text-mode command, invalid inside `$$...$$` math; it breaks
# pandoc's LaTeX -> Word (OMML) conversion (raw LaTeX in the doc). Strip for
# Word only. Emit as a standalone block (blank lines around `$$`) via cat so
# the multi-line equation is parsed as display math in every format.
.eq <- .assets$equation_latex
if (identical(fmt, "docx")) {
  .eq <- gsub("\\\\small", "", .eq)
  # Drop any now-blank lines: a blank line inside `$$...$$` terminates the math
  # block for pandoc and breaks the Word (OMML) conversion.
  .lines <- strsplit(.eq, "\n", fixed = TRUE)[[1]]
  .eq <- paste(.lines[nzchar(trimws(.lines))], collapse = "\n")
}
cat("\n$$\n", .eq, "\n$$\n\n", sep = "")
.eq_notation <- .assets$equation_notation
if (is.null(.eq_notation)) .eq_notation <- "linear"
if (identical(.eq_notation, "g")) {
  cat("See [Notation: Value Contributions](#sec-gnotation) at the end of this",
      "report for an explanation of the ${}^{f}g^{j}_{k}$ notation and its",
      "indexes.\n\n")
} else {
  cat("See [Notation: Value Contributions](#sec-gnotation) at the end of this",
      "report for an explanation of the value-contribution terms.\n\n")
}
```

## Coefficients

```{r coefficients}
knitr::kable(.assets$coef_df, digits = 4)
```

```{r enabled-interactions, results='asis', eval=.has_interactions}
# Enabled pairwise interactions (✓ = the x_i × x_j product term was included in
# the model matrix; the lasso may still have shrunk its coefficient to zero).
mat <- .imat
preds <- rownames(mat)
n <- length(preds)

display <- matrix("", n, n, dimnames = list(preds, preds))
for (i in seq_len(n)) {
  for (j in seq_len(n)) {
    display[i, j] <- if (i == j) "" else if (mat[i, j]) "✓" else "—"
  }
}

max_len <- if (n > 8) 12 else if (n > 5) 16 else 30
short <- ifelse(nchar(preds) > max_len,
                paste0(substr(preds, 1, max_len - 1), "…"), preds)
colnames(display) <- short
rownames(display) <- short

if (fmt == "latex") {
  cat("\n\\newpage\n\\begin{landscape}\n")
  cat("\\subsection{Enabled Interactions}\n\n")
  cat("\\footnotesize\n")
  print(knitr::kable(display, format = "latex", booktabs = TRUE))
  cat("\n\\normalsize\n")
  cat("\\end{landscape}\n\\newpage\n")
} else if (fmt == "docx") {
  esc_xml_ <- function(x) {
    x <- gsub("&", "&amp;", x, fixed = TRUE)
    x <- gsub("<", "&lt;",  x, fixed = TRUE)
    gsub(">", "&gt;",  x, fixed = TRUE)
  }
  rpr_ <- function(bold = FALSE) paste0(
    '<w:rPr><w:rFonts w:ascii="Arial Narrow" w:hAnsi="Arial Narrow" ',
    'w:cs="Arial Narrow"/>', if (bold) '<w:b/>' else '',
    '<w:sz w:val="16"/><w:szCs w:val="16"/></w:rPr>')
  cell_ <- function(txt, bold = FALSE) paste0(
    '<w:tc><w:tcPr><w:tcMar>',
    '<w:top w:w="20" w:type="dxa"/><w:bottom w:w="20" w:type="dxa"/>',
    '<w:left w:w="40" w:type="dxa"/><w:right w:w="40" w:type="dxa"/></w:tcMar>',
    '<w:vAlign w:val="center"/></w:tcPr>',
    '<w:p><w:pPr><w:jc w:val="center"/>', rpr_(bold), '</w:pPr>',
    '<w:r>', rpr_(bold), '<w:t xml:space="preserve">', esc_xml_(txt),
    '</w:t></w:r></w:p></w:tc>')
  brd_ <- paste0('<w:tblBorders>',
    paste(sprintf('<w:%s w:val="single" w:sz="4" w:space="0" w:color="999999"/>',
      c("top", "left", "bottom", "right", "insideH", "insideV")), collapse = ""),
    '</w:tblBorders>')
  header_row <- paste0('<w:tr>', cell_("", TRUE),
    paste(vapply(short, cell_, character(1), bold = TRUE), collapse = ""), '</w:tr>')
  body_rows <- paste(vapply(seq_len(n), function(i) {
    cells <- paste(vapply(seq_len(n), function(j) {
      if (i == j) cell_("") else if (mat[i, j]) cell_("✓") else cell_("—")
    }, character(1)), collapse = "")
    paste0('<w:tr>', cell_(short[i], TRUE), cells, '</w:tr>')
  }, character(1)), collapse = "")
  tbl_ <- paste0('<w:tbl><w:tblPr><w:tblW w:w="0" w:type="auto"/>', brd_,
    '</w:tblPr>', header_row, body_rows, '</w:tbl>')
  cat('\n\n```{=openxml}\n')
  cat('<w:p><w:pPr><w:sectPr><w:pgSz w:w="12240" w:h="15840"/>')
  cat('<w:pgMar w:top="1440" w:right="1440" w:bottom="1440" w:left="1440"/>')
  cat('</w:sectPr></w:pPr></w:p>\n')
  cat('```\n\n')
  cat("## Enabled Interactions\n\n")
  cat('```{=openxml}\n', tbl_, '\n```\n\n', sep = "")
  cat('```{=openxml}\n')
  cat('<w:p><w:pPr><w:sectPr><w:pgSz w:w="15840" w:h="12240" w:orient="landscape"/>')
  cat('<w:pgMar w:top="720" w:right="720" w:bottom="720" w:left="720"/>')
  cat('</w:sectPr></w:pPr></w:p>\n')
  cat('```\n\n')
} else {
  cat("\n## Enabled Interactions\n\n")
  cat('<style>\n')
  cat('.int-matrix-wrap { overflow-x: auto; margin: 1em 0; }\n')
  cat('.int-matrix { border-collapse: collapse; font-size: 0.8em; }\n')
  cat('.int-matrix th, .int-matrix td { border: 1px solid #ddd; padding: 3px 6px; text-align: center; }\n')
  cat('.int-matrix th.int-rotate { height: 120px; white-space: nowrap; vertical-align: bottom; }\n')
  cat('.int-matrix th.int-rotate > div { transform: rotate(-60deg); width: 1.5em; transform-origin: bottom left; margin-left: 50%; }\n')
  cat('.int-matrix td:first-child { text-align: left; font-weight: bold; white-space: nowrap; }\n')
  cat('.int-matrix .int-check { color: #27ae60; font-weight: bold; }\n')
  cat('.int-matrix .int-dash { color: #999; }\n')
  cat('</style>\n')
  cat('<div class="int-matrix-wrap">\n')
  cat('<table class="int-matrix">\n<thead><tr><th></th>')
  for (j in seq_len(n)) {
    cat(sprintf('<th class="int-rotate"><div>%s</div></th>', preds[j]))
  }
  cat('</tr></thead>\n<tbody>\n')
  for (i in seq_len(n)) {
    cat(sprintf('<tr><td>%s</td>', preds[i]))
    for (j in seq_len(n)) {
      if (i == j) {
        cat('<td></td>')
      } else if (mat[i, j]) {
        cat('<td class="int-check">✓</td>')
      } else {
        cat('<td class="int-dash">—</td>')
      }
    }
    cat('</tr>\n')
  }
  cat('</tbody></table>\n</div>\n')
}
```

## Variable Importance

```{r importance-plot, results='asis'}
include_plot("importance")
```

```{r importance-table}
knitr::kable(.assets$importance_df, digits = 4)
```

## Contributions

```{r contributions, results='asis'}
contrib_names <- .assets$contrib_names
for (vname in contrib_names) {
  safe_name <- gsub("[^a-zA-Z0-9_]", "_", vname)
  # Use a markdown heading with the variable name as inline code. Pandoc then
  # escapes underscores/specials correctly for every format. Emitting raw
  # LaTeX (\subsection{living_sqft}) breaks PDF with "Missing $ inserted".
  cat(sprintf("\n### `%s`\n\n", vname))
  is_int <- grepl(":", vname, fixed = TRUE)
  if (is_int) {
    # Interaction: scatter + heatmap + 3D persp
    include_plot(paste0("contrib_", safe_name, "_scatter"))
    include_plot(paste0("contrib_", safe_name, "_heatmap"))
    include_plot(paste0("contrib_", safe_name, "_3d"))
  } else {
    include_plot(paste0("contrib_", safe_name))
  }
  cat("\n\n")
}
```

## Correlation Matrix

```{r correlation, results='asis'}
include_plot("correlation")
```

## ANOVA Decomposition

```{r anova}
knitr::kable(.assets$anova_df, digits = 2)
```

## Diagnostics

### Coefficient Path

```{r coef-path, results='asis'}
include_plot("coef_path")
```

### Cross-Validation Error

```{r cv-error, results='asis'}
include_plot("cv_error")
```

### Actual vs Predicted

```{r actual-vs-predicted, results='asis'}
include_plot("actual_vs_predicted")
```

### Residuals vs Fitted

```{r residuals-vs-fitted, results='asis'}
include_plot("residuals_vs_fitted")
```

### Normal Q-Q Plot

```{r qq-plot, results='asis'}
include_plot("qq")
```

## Model Output

```{r model-output, comment=""}
cat("== glmnet Model ==\n\n")
print(.assets$model)
```

## Notation: Value Contributions {#sec-gnotation .unnumbered}

```{r g-legend, results='asis'}
.notation <- .assets$equation_notation
if (is.null(.notation)) .notation <- "linear"
cat(earthUI::g_function_legend_glmnet("markdown", .notation))
```

---

*Report generated by [glmnetUI](https://github.com/wcraytor/glmnetUI) on `r Sys.time()`*
