---
title: "The g-Function Notation for MARS Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The g-Function Notation for MARS Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

A MARS model (Multivariate Adaptive Regression Splines, implemented by the
`earth` package) produces a sum of a base constant plus additional terms, each
involving one to three variables. To analyze and visualize such a model, we
first split it into individual terms, then combine terms that share the same
set of variables into single interaction expressions called **g-functions**.

The `earthUI` package uses a compact g-function notation to organize model
terms into groups and to determine how each group should be graphed. This
vignette describes that notation and the reasoning behind it.

## Grouping Model Terms

Assume the maximum degree of interaction is 3. A fitted MARS model can be
partitioned into four groups of expressions:

- **Group 0**: the base constant.
- **Group 1**: first-degree expressions (single-variable terms).
- **Group 2**: second-degree interaction expressions (two-variable terms).
- **Group 3**: third-degree interaction expressions (three-variable terms).

Each group is graphed differently based on two factors:

1. The number of unique variables in the expression.
2. Whether any of those variables are **factor variables** (categorical).

Specifically:

- First-degree expressions (1 variable) are graphed in **2D**.
- Second-degree expressions (2 variables) are graphed in **3D**.
- Third-degree expressions (3 variables) are graphed as a set of **3D**
  graphs.
- If an expression contains a factor variable, each unique factor **reduces
  the graph's dimensionality by one** (e.g., a 3D graph becomes 2D if one
  variable is a factor).

## Indexing the g-Functions

Each function $g$ is indexed as ${}^{f_{j,k}}g^{j}_{k}$, where:

**The leading superscript** $f_{j,k}$ (before $g$):

- $j$ ranges from 0 to 3 and identifies the **group**:
    - $j = 0$: the base constant.
    - $j = 1$: first-degree expressions.
    - $j = 2$: second-degree expressions.
    - $j = 3$: third-degree expressions.
- $k$ is the index of a specific function **within** its group (e.g., $k = 1$
  for the first function in group $j$).
- The **value** of $f_{j,k}$ indicates the number of unique factor variables
  in the function:
    - $f_{j,k} = 0$: no factor variables.
    - $f_{j,k} > 0$: the count of factor variables (e.g., 1, 2, or 3).

**Trailing indices on** $g$:

- The superscript $j$ after $g$ repeats the group number for clarity.
- The subscript $k$ after $g$ matches the $k$ in $f_{j,k}$, identifying the
  function within its group.

While the indices of $f$ and $g$ align numerically, their roles differ:
$f_{j,k}$ counts factor variables (critical for graphing decisions), while $j$
and $k$ on $g$ define its group and position.

## Graphing Dimensions

The dimension $d$ of the graph for a function ${}^{f_{j,k}}g^{j}_{k}$ is:

$$d = j - f_{j,k}$$

where:

- $j$ is the number of unique variables (the group number).
- $f_{j,k}$ is the number of factor variables.
- $d$ is the resulting graph dimension.

**Examples:**

- $j = 3$ (3 variables), $f_{j,k} = 0$ (no factors): $d = 3$, requiring a
  **3D** graph.
- $j = 3$, $f_{j,k} = 1$ (one factor): $d = 2$, requiring a **2D** graph.
- $j = 2$, $f_{j,k} = 0$: $d = 2$, requiring a **2D** contour or surface.
- $j = 0$ (the constant): graphing is not needed.

When $d = 3$ (e.g., a third-degree expression with no factor variables),
visualizing the full function may require multiple 3D graphs, each fixing one
variable at different values to show a subset of the function's behavior.

## The Model Equation

The MARS model is broken into four groups of expressions. The indexing
${}^{f_{j,k}}g^{j}_{k}$ tells us:

- $j$: the group and number of variables.
- $f_{j,k}$: the number of factor variables.
- $d = j - f_{j,k}$: the graph dimension.

The estimated response $\hat{P}$ provided by the MARS model can be expressed
as:

$$
\hat{P} =
  {}^{f_{0,1}}g^{0}_{1}
  + \sum_{q=1}^{n_1} {}^{f_{1,q}}g^{1}_{q}(x^{*})
  + \sum_{r=1}^{n_2} {}^{f_{2,r}}g^{2}_{r}(x^{*}, y^{*})
  + \sum_{s=1}^{n_3} {}^{f_{3,s}}g^{3}_{s}(x^{*}, y^{*}, z^{*})
$$

where $n_1$, $n_2$, and $n_3$ are the number of first-, second-, and
third-degree g-functions, respectively.

## Number of Charts

The number of graphs required to visualize all g-functions in a model depends
on the variables, their interactions, and whether they are factor or non-factor
variables:

1. **Single-variable, non-factor** (${}^{0}g^{1}_{k}$): 1 line graph.
2. **Single-variable, factor** (${}^{1}g^{1}_{k}$): 1 bar graph combining all
   factor values.
3. **Base constant** (${}^{0}g^{0}_{1}$): no graph needed.
4. **2nd-degree, no factors** (${}^{0}g^{2}_{k}$): 1 surface or contour plot.
5. **2nd-degree, 1 factor** (${}^{1}g^{2}_{k}$): 1 graph per factor value
   present in the model.
6. **2nd-degree, 2 factors** (${}^{2}g^{2}_{k}$): 1 graph with a matrix of
   factor combinations.
7. **3rd-degree, no factors** (${}^{0}g^{3}_{k}$): multiple 3D graphs, each
   fixing one variable at different values.
8. **3rd-degree, 1 factor** (${}^{1}g^{3}_{k}$): 1 graph per factor value.
9. **3rd-degree, 2 factors** (${}^{2}g^{3}_{k}$): 1 graph per combination of
   factor values.
10. **3rd-degree, 3 factors** (${}^{3}g^{3}_{k}$): 1 graph per value of the
    factor with the fewest distinct values, using a color matrix for the other
    two.

For factor variables with many levels (e.g., an area identifier with 12
values interacting with living area), graphing can become complex. However,
`earth` rarely finds all combinations significant; typically only a few key
interactions warrant graphing, as identified by the model.

## Example

```{r}
library(earthUI)

result <- fit_earth(
  df = mtcars,
  target = "mpg",
  predictors = c("cyl", "disp", "hp", "wt", "qsec"),
  degree = 2
)

# List the g-functions and their indices
gf <- list_g_functions(result)
gf
```

The g-function table shows each group with its label, group index $j$
(`g_j`), position index $k$ (`g_k`), factor count $f$ (`g_f`), graph
dimension $d$, and number of terms. The `earthUI` package uses this structure
to automatically select the appropriate plot type for each g-function.

```{r fig.width=7, fig.height=4}
# Plot a first-degree g-function (2D line)
plot_g_function(result, 1)
```

## Reference

Craytor, W.B. (2025). *Real Estate Appraisal: A Regression-Centric
Approach*. Zenodo. DOI:
[10.5281/zenodo.14787917](https://doi.org/10.5281/zenodo.14787917)
